---
title: "DATA 607 Assignment 10A"
author: "Catherine Dube"
output: 
  html_document:
    code_folding: hide
---

```{r setup, include=FALSE}
# Load required libraries
if(!require(pacman)) install.packages("pacman")
pacman::p_load(tidyverse,
               stringr,
               tidytext,
               janeaustenr,
               textdata,
               wordcloud,
               reshape2,
               devtools,
               #vader,
               syuzhet,
               kableExtra,
               scales
               )
```

# Intro
In this assignment, we've been asked to start with the example code from [Text Mining with R, Chapter 2 Sentiment analysis with tidy data](https://www.tidytextmining.com/sentiment.html) (Silge and Robinson).

I will then be extending the analysis to: 

* Work with a different corpus: Harry Potter
* Incorporate at an additional sentiment lexicon: Loughran
  * Loughran is said to be good for fictional works, so it should be a good fit here.

# Sections {.tabset}

## Example Code 
This section is entirely sourced from [Text Mining with R, Chapter 2 Sentiment analysis with tidy data](https://www.tidytextmining.com/sentiment.html), by Silge and Robinson.

### Loading in Jane Austen Book Data

```{r given-example} 
tidy_books <- austen_books() %>%
  group_by(book) %>%
  mutate(
    linenumber = row_number(),
    chapter = cumsum(str_detect(text, 
                                regex("^chapter [\\divxlc]", 
                                      ignore_case = TRUE)))) %>%
  ungroup() %>%
  unnest_tokens(word, text)
```


### Initial Sentiment Analysis

```{r sentiment-example}
nrc_joy <- get_sentiments("nrc") %>% 
  filter(sentiment == "joy")

tidy_books %>%
  filter(book == "Emma") %>%
  inner_join(nrc_joy) %>%
  count(word, sort = TRUE)
```
```{r sentiment-example-2, warning = FALSE}
jane_austen_sentiment <- tidy_books %>%
  inner_join(get_sentiments("bing")) %>%
  count(book, index = linenumber %/% 80, sentiment) %>%
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>% 
  mutate(sentiment = positive - negative)

ggplot(jane_austen_sentiment, aes(index, sentiment, fill = book)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~book, ncol = 2, scales = "free_x")
```

### Comparing Sentiment Dictionaries

```{r sentiment-dictionary, warning = FALSE}
pride_prejudice <- tidy_books %>% 
  filter(book == "Pride & Prejudice")

afinn <- pride_prejudice %>% 
  inner_join(get_sentiments("afinn")) %>% 
  group_by(index = linenumber %/% 80) %>% 
  summarise(sentiment = sum(value)) %>% 
  mutate(method = "AFINN")

bing_and_nrc <- bind_rows(
  pride_prejudice %>% 
    inner_join(get_sentiments("bing")) %>%
    mutate(method = "Bing et al."),
  pride_prejudice %>% 
    inner_join(get_sentiments("nrc") %>% 
                 filter(sentiment %in% c("positive", 
                                         "negative"))
    ) %>%
    mutate(method = "NRC")) %>%
  count(method, index = linenumber %/% 80, sentiment) %>%
  pivot_wider(names_from = sentiment,
              values_from = n,
              values_fill = 0) %>% 
  mutate(sentiment = positive - negative)

bind_rows(afinn, 
          bing_and_nrc) %>%
  ggplot(aes(index, sentiment, fill = method)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~method, ncol = 1, scales = "free_y")
```

### Most common positive & negative words

```{r example-compare}
bing_word_counts <- tidy_books %>%
  inner_join(get_sentiments("bing")) %>%
  count(word, sentiment, sort = TRUE) %>%
  ungroup()

bing_word_counts %>%
  group_by(sentiment) %>%
  slice_max(n, n = 10) %>% 
  ungroup() %>%
  mutate(word = reorder(word, n)) %>%
  ggplot(aes(n, word, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~sentiment, scales = "free_y") +
  labs(x = "Contribution to sentiment",
       y = NULL)
```

```{r example-stop-words}
custom_stop_words <- bind_rows(tibble(word = c("miss"),  
                                      lexicon = c("custom")), 
                               stop_words)
```

### Wordcloud
```{r example-cloud, warning = FALSE}
tidy_books %>%
  anti_join(stop_words) %>%
  count(word) %>%
  with(wordcloud(word, n, max.words = 100))
```

```{r example-wordcloud-positive-neg, warning = FALSE}
tidy_books %>%
  inner_join(get_sentiments("bing")) %>%
  count(word, sentiment, sort = TRUE) %>%
  acast(word ~ sentiment, value.var = "n", fill = 0) %>%
  comparison.cloud(colors = c("gray20", "gray80"),
                   max.words = 100)
```

## Harry Potter Code

The lovely Data Scientist, Bradley Boehmke, maintains [this GitHub Repo](https://github.com/bradleyboehmke/harrypotter). It is, in his own words "An R Package for J.K. Rowling's Harry Potter Series". 

### Load in Harry Potter Data

```{r load-hp, warning = FALSE}
devtools::install_github("bradleyboehmke/harrypotter") 
library(harrypotter)
```

### Initial Sentiment Analysis 

To start off, I'm looking at the positive/negative sentiment per book using the bing lexicon.

```{r sentiment-stuff}
# Combine all books 
hp_books <- bind_rows(
  tibble(text = philosophers_stone)   %>% mutate(book = "Philosopher's Stone"),
  tibble(text = chamber_of_secrets)   %>% mutate(book = "Chamber of Secrets"),
  tibble(text = prisoner_of_azkaban)  %>% mutate(book = "Prisoner of Azkaban"),
  tibble(text = goblet_of_fire)       %>% mutate(book = "Goblet of Fire"),
  tibble(text = order_of_the_phoenix) %>% mutate(book = "Order of the Phoenix"),
  tibble(text = half_blood_prince)    %>% mutate(book = "Half-Blood Prince"),
  tibble(text = deathly_hallows)      %>% mutate(book = "Deathly Hallows")
)

# Token
hp_tidy <- hp_books %>%
  unnest_tokens(word, text) %>%
  group_by(book) %>%
  mutate(linenumber = row_number()) %>%       # number each token (word) within book
  ungroup()
```

```{r bing-stuff}
# Cleaning bing because envious was in there twice - as both positive and negative
cleaned_bing <- get_sentiments("bing") %>%
  filter(!((word == 'enviousness') & (sentiment == 'positive'))) %>%
  filter(!((word == 'envious') & (sentiment == 'positive'))) %>%
  filter(!((word == 'enviously') & (sentiment == 'positive')))

# Compute sentiment per 300-line chunk 
hp_bing_sentiment <- hp_tidy %>% 
  inner_join(cleaned_bing, by = "word") %>% 
  mutate(index = linenumber %/% 300) %>% # chunk every 80 words per book 
  count(book, index, sentiment) %>% 
  pivot_wider(names_from = sentiment, values_from = n, values_fill = 0) %>% 
  mutate(sentiment = positive - negative)

# Wanted to have global index to treat harry potter as continuous book
offsets <- hp_bing_sentiment %>%
  group_by(book) %>%
  summarize(book_offset = max(index) + 1) %>%
  mutate(book = factor(book, levels = c(
    "Philosopher's Stone",
    "Chamber of Secrets",
    "Prisoner of Azkaban",
    "Goblet of Fire",
    "Order of the Phoenix",
    "Half-Blood Prince",
    "Deathly Hallows"
  ))) %>%
  arrange(book) %>%
  mutate(cum_offset = cumsum(lag(book_offset, default = 0)))

# Join offsets back in
hp_bing_sentiment <- hp_bing_sentiment %>%
  left_join(offsets %>% select(book, cum_offset), by = "book") %>%
  mutate(global_index = index + cum_offset)
```

From these below views, it looks like the sentiment leans slightly negative, though not by much. 

```{r bing-plots, fig.width=12, fig.height=10}
# Plot Bing Sentiment Analysis Per Book
ggplot(hp_bing_sentiment, aes(index, sentiment, fill = book)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~book, ncol = 2, scales = "free_x")

# Plot Bing Sentiment Analysis as one continuous book
ggplot(hp_bing_sentiment, aes(global_index, sentiment, fill = book)) +
  geom_col(show.legend = TRUE, width = 1) +
  labs(
    title = "Harry Potter Series: Continuous Sentiment (Bing Lexicon)",
    x = "Progression Across Series (per 300-word chunk)",
    y = "Net Sentiment (Positive - Negative)",
    fill = "Book"
  ) +
  theme_minimal(base_size = 14)

```

```{r nrc-stuff}
# NRC
# anger, anticipation, disgust, fear, joy, negative, positive, sadness, surprise, trust
hp_nrc_sentiment <- suppressWarnings(
  hp_tidy %>% 
  inner_join(get_sentiments("nrc"), by = "word") %>%
  mutate(index = linenumber %/% 300) %>%
  count(book, index, sentiment) %>%
  pivot_wider(
    names_from = sentiment,
    values_from = n,
    values_fill = 0
  ) %>%
  mutate(sentiment = positive - negative)
)

# Join offsets 
hp_nrc_sentiment <- hp_nrc_sentiment %>%
  left_join(offsets %>% select(book, cum_offset), by = "book") %>%
  mutate(global_index = index + cum_offset)
```

```{r afinn}
# AFINN 
# values[-5,5] 
hp_afinn_sentiment <- hp_tidy %>%
  inner_join(get_sentiments("afinn"), by = "word") %>%
  mutate(index = linenumber %/% 300) %>%
  group_by(book, index) %>%
  summarize(sentiment = sum(value, na.rm = TRUE), .groups = "drop")

# Join offsets
hp_afinn_sentiment <- hp_afinn_sentiment %>%
  left_join(offsets %>% select(book, cum_offset), by = "book") %>%
  mutate(global_index = index + cum_offset)

```

```{r loughran-stuff}
# Loughran 
# constraining, litigious, negative, positive, superfluous, uncertainty
hp_loughran_sentiment <- suppressWarnings(
  hp_tidy %>%
    inner_join(get_sentiments("loughran"), by = "word") %>%
    mutate(index = linenumber %/% 300) %>%
    count(book, index, sentiment) %>%
    pivot_wider(
      names_from = sentiment,
      values_from = n,
      values_fill = 0
    ) %>%
    mutate(sentiment = positive - negative)
)

# Join offsets
hp_loughran_sentiment <- hp_loughran_sentiment %>%
  left_join(offsets %>% select(book, cum_offset), by = "book") %>%
  mutate(global_index = index + cum_offset)
```

Let's take a similar look, but use multiple different lexicons to see if the positive/negative distribution looks roughly the same or not.

```{r compare-methods, fig.width=12, fig.height=10}
# Combine all sentiment methods  
hp_all_sentiments <- bind_rows(
  hp_bing_sentiment  %>% mutate(method = "Bing"),
  hp_nrc_sentiment   %>% mutate(method = "NRC"),
  hp_afinn_sentiment %>% mutate(method = "AFINN"),
  hp_loughran_sentiment %>% mutate(method = "Loughran")
)

# Plotting
ggplot(hp_all_sentiments, aes(global_index, sentiment, fill = book)) +
  geom_col(show.legend = FALSE, width = 1) +
  facet_wrap(~method, ncol = 1, scales = "free_y") +
  labs(
    title = "Harry Potter: Sentiment Comparison Across Each Lexicon",
    x = "Progression Across Series (per 300-word chunk)",
    y = "Net Sentiment (Positive - Negative)"
  ) +
  theme_minimal(base_size = 14) +
  theme(
    strip.text = element_text(face = "bold", size = 13),
    plot.title = element_text(face = "bold")
  )

```

This has all been visually pretty and interesting so far, but I feel like some of the definitive values are getting lost. I think that the sentiment is looking more negative regardless of lexicon, but here, I'm going to take a look at it from a summary table view.

```{r more-compare-methods}
hp_sentiment_summary <- hp_all_sentiments %>%
  group_by(method) %>%
  summarize(
    total_chunks = n(),
    positive_chunks = sum(sentiment > 0, na.rm = TRUE),
    negative_chunks = sum(sentiment < 0, na.rm = TRUE),

    # 🆕 total words counted across all chunks (proxy for tone intensity)
    positive_words = sum(pmax(sentiment, 0), na.rm = TRUE),
    negative_words = sum(pmax(-sentiment, 0), na.rm = TRUE),

    percent_positive = positive_chunks / total_chunks,
    percent_negative = negative_chunks / total_chunks,
    .groups = "drop"
  ) %>%
  mutate(
    percent_positive = percent(percent_positive, accuracy = 0.1),
    percent_negative = percent(percent_negative, accuracy = 0.1)
  ) %>%
  select(
    method,
    total_chunks,
    positive_chunks,
    negative_chunks,
    positive_words,
    negative_words,
    percent_positive,
    percent_negative
  )

hp_sentiment_summary %>%
  kable("html", caption = "Harry Potter Sentiment Summary by Lexicon") %>%
  kable_styling(
    full_width = FALSE,
    position = "center",
    bootstrap_options = c("striped", "hover")
  )
```



Now from the table above, I'm seeing that using the AFINN lexicon, we're flagging the highest volume of words: `r format(hp_sentiment_summary[hp_sentiment_summary$method == "AFINN", "positive_words"] + hp_sentiment_summary[hp_sentiment_summary$method == "AFINN", "negative_words"], scientific=FALSE)` words total. AFINN also shows the most positive sentiment out of analysis with any lexicon, yet it's still only `r hp_sentiment_summary[hp_sentiment_summary$method == "AFINN", "percent_positive"]`. The sentiment in Harry Potter, regardless of lexicon, is majority negative.

Now that I've determine I like looking at the data from this percent positive and negative perspective, I want to look at this broken down per book.

```{r book-pos-neg, fig.width=8, fig.height=6} 
#book summary
hp_book_summary <- hp_all_sentiments %>%
  group_by(method, book) %>%
  summarize(
    total_chunks = n(),
    positive_chunks = sum(sentiment > 0, na.rm = TRUE),
    negative_chunks = sum(sentiment < 0, na.rm = TRUE),
    percent_positive = positive_chunks / total_chunks,
    percent_negative = negative_chunks / total_chunks,
    .groups = "drop"
  ) %>%
  pivot_longer(
    cols = c(percent_positive, percent_negative),
    names_to = "polarity",
    values_to = "percent"
  ) %>%
  mutate(
    polarity = recode(polarity,
                      "percent_positive" = "Positive",
                      "percent_negative" = "Negative")
  )
# Book order
book_order <- c(
  "Philosopher's Stone",
  "Chamber of Secrets",
  "Prisoner of Azkaban",
  "Goblet of Fire",
  "Order of the Phoenix",
  "Half-Blood Prince",
  "Deathly Hallows"
)

# Re order the summary
hp_book_summary <- hp_book_summary %>%
  mutate(book = factor(book, levels = book_order))

# plot pos/neg by book
ggplot(hp_book_summary, aes(x = book, y = percent, fill = polarity)) +
  geom_col(position = "dodge") +
  facet_wrap(~method, ncol = 1) +
  scale_y_continuous(labels = percent_format(accuracy = 1)) +
  scale_fill_manual(values = c("Positive" = "#2ca02c", "Negative" = "#d62728")) +
  labs(
    title = "Percent Positive vs. Negative Chunks per Book",
    x = "Book",
    y = "Percent of Chunks",
    fill = "Polarity"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold", size = 12),
    axis.text.x = element_text(angle = 45, hjust = 1),
    plot.title = element_text(face = "bold", size = 14)
  )

```

A couple interesting things of note from that! 

* Deathly Hallows consistently seems to be the book that has the most negative sentiment in it.  
* We already knew that AFINN identified the most positive sentiment. But it's also the only lexicon's analysis that results in some books showing more positive sentiment than negative sentiment: namely The Philosopher's Stone and Half-Blood Prince.

One last thing comparing the different lexicons, I know that some lexicons bucket into categories beyond just "positive" and "negative". So I wanted to look at that distribution across the series. I am especially interested in the findings from Loughran here, because it's a lexicon well-suited for fiction.

```{r sentiments-per-lexicon} 

# Bing
suppressWarnings(
bing_counts <- hp_tidy %>%
  inner_join(get_sentiments("bing"), by = "word") %>%
  count(sentiment) %>%
  mutate(method = "Bing")
)
# NRC 
suppressWarnings(
nrc_counts <- hp_tidy %>%
  inner_join(get_sentiments("nrc"), by = "word") %>%
  count(sentiment) %>%
  mutate(method = "NRC")
)

# AFINN 
# numeric scores: group into positive / negative / neutral
suppressWarnings(
afinn_counts <- hp_tidy %>%
  inner_join(get_sentiments("afinn"), by = "word") %>%
  mutate(sentiment = case_when(
    value > 0 ~ "positive",
    value < 0 ~ "negative",
    TRUE ~ "neutral"
  )) %>%
  count(sentiment) %>%
  mutate(method = "AFINN")
)

# Loughran 
suppressWarnings(
loughran_counts <- hp_tidy %>%
  inner_join(get_sentiments("loughran"), by = "word") %>%
  count(sentiment) %>%
  mutate(method = "Loughran")
)

# Combine all
all_lexicon_counts <- bind_rows(bing_counts, nrc_counts, afinn_counts, loughran_counts)

ggplot(all_lexicon_counts, aes(x = reorder(sentiment, n), y = n, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  coord_flip() +
  facet_wrap(~method, scales = "free_y", ncol = 2) +
  scale_y_continuous(labels = scales::comma) +
  labs(
    title = "Distribution of Sentiment Categories Across Lexicons",
    x = "Sentiment Category",
    y = "Word Count"
  ) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold", size = 13),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

```

This was neat, and while it reiterates a lot that we already know, we are now able to see that there is a lot of "uncertainty", "sadness" and "anger" reflected in the language use in the books - outside of just "positive" and "negative" sentiment.

### Extra AFINN Sentiment Analysis 

Opting to dig into some more AFINN analysis because it is compartmentalizing the highest number of words (as we saw earlier).

```{r afinn-analysis-pos-neg}

# Join AFINN to your tidy Harry Potter tokens
hp_afinn_words <- hp_tidy %>%
  inner_join(get_sentiments("afinn"), by = "word")

# Compute total contribution per word across all books
afinn_word_counts <- hp_afinn_words %>%
  group_by(word) %>%
  summarize(
    total_value = sum(value),          # net sentiment (signed sum)
    abs_contribution = sum(abs(value)) # magnitude for ranking
  ) %>%
  ungroup() %>%
  mutate(sentiment = ifelse(total_value > 0, "positive", "negative"))

# Select top 10 most common (strongest) positive & negative words
top_afinn_words <- afinn_word_counts %>%
  group_by(sentiment) %>%
  slice_max(abs_contribution, n = 10) %>%
  ungroup() %>%
  mutate(word = reorder_within(word, abs_contribution, sentiment))

# Plot
ggplot(top_afinn_words, aes(x = abs_contribution, y = word, fill = sentiment)) +
  geom_col(show.legend = FALSE) +
  facet_wrap(~sentiment, scales = "free_y") +
  scale_y_reordered() +
  labs(
    title = "Most Common Positive & Negative Words (using AFINN Lexicon)",
    x = "Weighted Contribution to Sentiment",
    y = NULL
  ) +
  theme_minimal(base_size = 13) +
  theme(
    strip.text = element_text(face = "bold", size = 13),
    plot.title = element_text(face = "bold", hjust = 0.5)
  )

```

### Word Cloud

Finally I wanted to make some word clouds to get an idea of what words appear the most frequently in each book. (The word clouds live up to their expectations, the most common word in Harry Potter is consistently "Harry". )

```{r word-cloud, fig.width=12, fig.height=12}
# remove stop words 
hp_word_freq <- hp_tidy %>%
  anti_join(stop_words, by = "word") %>%
  count(book, word, sort = TRUE)

# Set up the plotting area
#par(mfrow = c(3, 3))  # 7 books → fits nicely into a 3x3 grid
#par(mfrow = c(2, 4), mar = c(0, 0, 2, 0))  # mar=c(bottom,left,top,right)
par(
  mfrow = c(3, 3),
  mar = c(0, 0, 2, 0),       # inner margins (per plot)
  oma = c(0, 0, 0, 0)        # outer margins (around all plots)
)

suppressWarnings(
  book_order %>%
    purrr::walk(function(bk) {
      book_df <- filter(hp_word_freq, book == bk)
  
      # Dense and high-contrast word cloud
      wordcloud(
        words = book_df$word,
        freq = book_df$n,
        max.words = 100,
        random.order = FALSE,
        scale = c(4.5, 0.9),       # bigger range = fills more space
        rot.per = 0.05,          # fewer rotated words (less wasted space)
        colors = brewer.pal(8, "Dark2"),
        use.r.layout = FALSE     # pack words tighter (uses C routine)
      )
  
      title(bk, cex.main = 1, font.main = 1.5, line = 1)
    })
)
```


